---
title: 'Class 19: Genomics'
author: 'Shitian Li (PID: A13294481)'
date: "12/6/2021"
output: pdf_document
---

```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
```

```{r}
data <- read.table('rs8067378_ENSG00000172057.6.txt')
AA <- filter(data, geno=='A/A')
AG <- filter(data, geno=='A/G')
GG <- filter(data, geno=='G/G')
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
nrow(AA)
nrow(AG)
nrow(GG)
median(AA$exp)
median(AG$exp)
median(GG$exp)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
ggplot(data, aes(x=as.factor(geno), y=exp)) + 
  geom_boxplot() + 
  labs(x="Genotype", y="Expression")
```

It seems like expression of G/G is much lower than that of A/A. This SNP, indeed, affects the expression of ORMDL3. 
