---
title: "Class 15"
author: 'Shitian Li (PID: A13294481)'
date: "11/17/2021"
output: pdf_document
---

# background

Today we examine a published RNA-seqexperiment where airway smooth muscle cells were treated with 

```{r include=FALSE}
library(dplyr)
library(ggplot2)
```

We need 2 things:  
-1: count data   
-2: colData (the metadata that tells us about the deign of the experiment)  

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

Side note:  
Let's check the correspondence of the metadata and count data setup. 

```{r}
all(metadata$id == colnames(counts))
```

# Compare control to treated

First, we need to access all the countrol columns in our counts data. 

```{r}
control.inds <- metadata$dex == "control"
control.ids <- metadata[control.inds,]$id
```

Use these ids to access just the control columns f our `counts` data

```{r}
control.mean <- rowMeans(counts[ ,control.ids])
head(counts[ , control.ids])
```

Now do same for treat: 

```{r}
treat.inds <- !control.inds
treat.ids <- metadata[treat.inds,]$id
treat.mean <- rowMeans(counts[ ,treat.ids])
head(counts[ , treat.ids])
```

For book keeping, let's combine these into a new dataset. 

```{r}
meancounts <- data.frame(control.mean, treat.mean)
```

There are `r nrow(counts)` rows/genes in this dataset. 

```{r}
nrow(counts)
nrow(meancounts)
```

# Compare the control and treated: 

Let's do a quick plot: 

```{r}
plot(meancounts$control.mean, meancounts$treat.mean)
```

Let's put on a log scale: 

```{r}
plot(meancounts$control.mean, meancounts$treat.mean, log="xy")
```

We often use log transformations as they make life much nicer in this world...

```{r}
log2(20/20)
```

```{r}
log2(40/20)
```

```{r}
log2(10/20)
```

```{r}
log2(80/20)
```

```{r}
meancounts$log2fc <- log2(meancounts[,"treat.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

Remove those with inf and NaN. 

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

We now have   `r nrow(mycounts)` gene remaining: 

```{r}
nrow(mycounts)
```

How many of these genes are up regulated at the log2 fold-change threshold of +2 or greater? 

```{r}
sum(mycounts$log2fc > +2)
```

What percentage is this? 

```{r}
round((sum(mycounts$log2fc > +2) / nrow(mycounts)) * 100, 2)
```

```{r}
sum(mycounts$log2fc < -2)
```

# DEseq2

```{r include=F}
library(DESeq2)
```

```{r}
citation("DESeq2")
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
dds
```
 
Run the DESeq analysis: 

```{r}
dds <- DESeq(dds)
```

Get the results: 

```{r}
res <- results(dds)
res
```

Let's take a look at the summary of results: 

```{r}
summary(res)
```

Change the p-value to 0.05 from the 0.1 default: 

```{r}
res05 <- results(dds, alpha=0.05)
summary(res05)
```

# A volcano plot

This is a very common data viz of this type of data that does not really look like a volcano. 

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
      xlab="Log2(FoldChange)",
      ylab="-Log(P-value)")
```

```{r}
plot( res$log2FoldChange,  -log(res$padj), 
 ylab="-Log(P-value)", xlab="Log2(FoldChange)")

# Add some cut-off lines
abline(v=c(-2,2), col="darkgray", lty=2)
abline(h=-log(0.05), col="darkgray", lty=2)
```

```{r}
# Setup our custom point color vector 
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

# Volcano plot with custom colors 
plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

# Cut-off lines
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r include=FALSE}
library(EnhancedVolcano)
```

# Now let's add in some labels: 

```{r include=FALSE}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

Store the data: 

```{r}
write.csv(res[ord,], "deseq_results.csv")
```

And then do enhanced volcano plot: 

```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```






