---
title: "Mini Project"
author: 'Shitian Li (PID: A13294481)'
date: "10/27/2021"
output: pdf_document
---

## Data Import

```{r include=FALSE}
library(dplyr)
```

```{r Import}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)

# Examine if the data is imported correctly: 
head(wisc.df)
```

Seems like our data is fine. Now let's clean up a little bit: 

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
```

Also want to save the diagnostics to compare later: 

```{r}
# Create diagnosis vector for later 
diagnosis <- as.factor(wisc.df[,1])
```


> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(diagnosis == "M")
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
end.with.mean <- wisc.data %>%
 select(ends_with("_mean"))
ncol(end.with.mean)
```

## PCA

Now, let's check if we need to scale the data by checking the mean and variances: 

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

Seems like we have a large SD and difference among the means. Now let's execute the PCA with scale turned on: 

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp( wisc.data, scale. = T )
```  

and now let's inspect the data: 

```{r}
# Look at summary of results
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?  

44.27%

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7

## Interpreting PCA results

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

There are too many words and too many data inter-tangled together in this plot, which makes it hard to understand. We need to examine less PCs. 

Now let's plot PC1 vs. PC2: 

```{r}
# Scatter plot observations by components 1 and 2
plot( wisc.pr$x[,1], wisc.pr$x[,2] , col = diagnosis , 
     xlab = "PC1", ylab = "PC2")
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[, 1], wisc.pr$x[, 3], col = diagnosis, 
     xlab = "PC1", ylab = "PC3")
```
It is note worthy that we can see a significant difference along PC1 between malignant and normal cells. However, we cannot get much information about PC2 and PC3. 

Now let's explore ggplot2: 

```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)

# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
  geom_point()
```

Now let's analyze the variance: 

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
# install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```


## Communicating PCA results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation['concave.points_mean',1]
```

> Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

```{r}
length(pve) - sum(cumsum(pve) > 0.8) + 1
```

## Hierarchical clustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
head(data.dist)
```

```{r}
wisc.hclust <- hclust(data.dist, method="complete")
```

```{r}
plot(wisc.hclust)
abline(19, 0, col="red", lty=2)
```

> Q11. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

Seems like 19 would work. 

## Cut tree: 

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
```

```{r}
table(wisc.hclust.clusters, diagnosis)
```

> Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

From the table printed below, it seems like 4 or 5 works well. Other numbers are not as good. 

```{r}
for (i in 2:10) {
  wisc.hclust.clusters.new <- cutree(wisc.hclust, k=i)
  print(paste("cut into", i))
  print(table(wisc.hclust.clusters.new, diagnosis))
}

```

## Different Methods

> Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

I tried multiple methods from below. Seems like **ward.D** and **ward.D2** works pretty well if we want two groups, one malignant and one normal. There is a clear difference between two groups for these two methods. 

```{r}
methods <- c("ward.D", "ward.D2", "single", "complete", "average", "mcquitty", "median", "centroid")
wisc.new.hclust <- hclust(data.dist, method=methods[2])
plot(wisc.new.hclust)
```


## k-means

```{r}
wisc.km <- kmeans(wisc.data, centers= 2, nstart= 100)
table(wisc.km$cluster, diagnosis)
```

> Q14. How well does k-means separate the two diagnoses? How does it compare to your hclust results?

I think the result is acceptable. It's close to what we got in hierarchical clustering, but not as good. 

```{r}
table(wisc.hclust.clusters, diagnosis)
```

## Combining Methods

Using 90% variance, that is, 7 PCs, we want to hierarchical clustering the PCAs using `method="ward.D2"`

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[,1:7]), method="ward.D2")
plot(wisc.pr.hclust)
```

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```


```{r}
table(grps, diagnosis)
```

```{r}
g <- as.factor(grps)
levels(g)
g <- relevel(g,2)
levels(g)
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
plot(wisc.pr$x[,1:2], col=diagnosis)
```

> Q15. How well does the newly created model with four clusters separate out the two diagnoses?

Seems to be working well. Better than without the PCA before. 

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
table(wisc.pr.hclust.clusters, diagnosis)
```

> Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses.

As shown below, they're not as good as PCA+Hclust together. 

```{r}
table(wisc.km$cluster, diagnosis)
table(wisc.hclust.clusters, diagnosis)
```


> Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

From below, seems combined method is the best for both sensitivity and specificity. 

```{r}
kmeans.ss <- c(130/(130+82), 356/(356+82))
hclust.ss <- c(165/(165+40), 343/(343+40))
combine.ss <- c(188/(188+24), 329/(329+24))
ss <- data.frame(kmeans.ss, hclust.ss, combine.ss, row.names = c("Sensitivity", "Specificity"))
print(ss)
```


## Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q18. Which of these new patients should we prioritize for follow up based on your results?

From the above result, seems like patient 2 should be prioritized. 
