---
title: 'Class 6: R Functions'
author: 'Shitian Li (PID: A13294481)'
date: "10/15/2021"
output: github_document
---

## Quick Rmarkdown intro

We can write text, of course, just like any file. We can style text to be **bold** or *italic*. 

Do: 

- this
- and that
- and another thing

This is more text  
and this is a new line.

---

We can include some code:

```{r}
plot(1:10)
```

```{r}
# This is a comment and will not be passed to R
plot(cars)
```


## Time to write a function

> **Q1**. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped. Your final function should be adquately explained with code comments and be able to work on an example class gradebook such as this one in CSV format: “https://tinyurl.com/gradeinput” [3pts]

```{r}
# import some example vectors to test
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90) 
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80) 
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
```

First, want to find lowest score. We can use **min()** and **which.min()** functions to find them. 
```{r}
which.min(student1)
```

We can use **-** to delte a value
```{r}
student1[-which.min(student1)]
```

Now call the **mean()** function: 
```{r}
mean(student1[-which.min(student1)])
```

This does not work on student 2 because of the *NA* value: 
```{r}
mean(student2[-which.min(student2)])
```
OK, let's change those NA values into 0
```{r}
student2[is.na(student2)] <- 0
mean(student2[-which.min(student2)])
```

OK. I'll skip Barry's instructions and directly work on student3 haha. 
```{r}
NA_position <- which(is.na(student3))
student3[NA_position] <- 0
mean(student3[-which.min(student3)])
```

```{r}
x <- 1:5
x
x[2] <- 100
x
x[3] <- 200
x
```

Great! we got it. Now let's simplify and make it as simple as possible. 

We can make variable names more clear

```{r}
x <- student3
x[ is.na(x)] = 0
mean(x[-which.min(x)])

```

What if the data is wrong: 
```{r}
student4 <- c(100, NA, 90, "90", 90, 90, 97, 80)
x <- as.numeric(student4)
x[ is.na(x)] = 0
mean(x[-which.min(x)])
```



OK, finally we're writing the function: 

All functions have at least 3 things:  
A name, input args, and a body.
```{r}
grade <- function(x) {
  x <- as.numeric(x)
  x[is.na(x)] = 0
  return(mean(x[-which.min(x)]))
}
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90) 
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80) 
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
grade(student1)
grade(student2)
grade(student3)
```

## Now grade a whole class

First, read the gradebook for the class. 

```{r}
gradebook <- "https://tinyurl.com/gradeinput"
scores <- read.csv(gradebook, row.names=1)
```


We're going to use **apply()** function to grade all the students with the **grade()** function
```{r}
ans <- apply(scores, 1, grade)
ans
```

> **Q2**. Who is the top scoring student? 

```{r}
which.max(ans)
```

> **Q3**. Which homework hardest? 

Here I'll use the **apply()** function again but look at columns. 

Maybe we can ignore the NA values with na.rm=T? 
```{r}
apply(scores, 2, mean, na.rm=T)
```

Let's replace or mask NA values to zero
```{r}
mask <- scores
mask[is.na(mask)]=0
hw_avg <- apply(mask, 2, mean)
hw_avg
which.min(apply(mask, 2, mean))
```

> **Q4**. Which homework is the most representative?

Here we'll use the **cor()** function here
```{r}
ans <- apply(scores, 1, grade)
hw_cor <- apply(mask, 2, cor, ans)
hw_cor
which.max(hw_cor)
```

```{r}
student <- c(100, 100, 100, 90)
grade2 <- function(x, drop.lowest=T) {
  x <- as.numeric(x)
  x[is.na(x)] = 0
  if (drop.lowest) {
    mean(x[-which.min(x)])
  } else {
    mean(x)
  }
}
grade2(student, F)
```


