---
title: "Class 10: Halloween Project"
author: 'Shitian Li (PID: A13294481)'
date: "10/31/2021"
output: github_document
always_allow_html: true
---

## Import data
```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

### What does this dataset contain? 

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

## Favorite candy? 

```{r}
candy["Twix", ]$winpercent
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Haribo Gold Bears", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

## Skim: 

```{r}
# install.packages("skimr")
library("skimr")
skim(candy)
```

Use `skim()` and answer the following:

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Winpercent is in a different scale to majority since it has a very high STD. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

0: does not contain chocolate
1: contains chocolate

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```

> Q9. Is the distribution of winpercent values symmetrical?

Yes. 

> Q10. Is the center of the distribution above or below 50%?

Below. 
```{r}
median(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
candy.choco.type <- candy$winpercent[as.logical(candy$chocolate)]
candy.fruit.type <- candy$winpercent[as.logical(candy$fruity)]

t.test(candy.choco.type, candy.fruit.type)
```
As a result, we see that chocolate is higher ranked than fruity type.

> Q12. Is this difference statistically significant?

Yes, with *p = 2.871e-08*. 


## Overall candy rankings

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>%
  arrange(winpercent) %>%
  head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>%
  arrange(desc(winpercent)) %>%
  head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=10}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

### Set up color

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r, fig.height=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

sixlets

> Q18. What is the best ranked fruity candy?

starbusts

## Pricepercent

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip
```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

> *opitional* Q21. Make a barplot again with `geom_col()` this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping `geom_col()` for `geom_point()` + `geom_segment()`.

```{r, fig.height = 10}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

## Correlation

```{r}
# install.packages("corrplot")
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and Fruity

> Q23. Similarly, what two variables are most positively correlated?

Winpercent and Chocolate

## PCA

```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])

p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r include=FALSE}
library(plotly)
```

```{r}
ggplotly(p)
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, pluribus, and hard. Yes, as fruity candies are usually hard nad pluribus, and not the other components.
